---
title: "Tycho"
author: "Tushal Kakadiya"
date: '2022-06-21'
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
header-includes:
- \usepackage{titling}
- \pretitle{\begin{center} \includegraphics[width=5in,height=5in]{THD-Logo_grau.png}\LARGE\\}
- \posttitle{\end{center}}
---


# 1.Introduction

When it comes to interpreting the world and the enormous amount of data is producing on a daily basis, Data Visualization becomes the most desirable way. Rather than screening huge Excel sheets, it is always better to visualize that data through charts and graphs, to gain meaningful insights. 


# 2.Objectives

We are trying to study the nature of spread of different types of diseases in the all states of United States.
For that we will be using the data provided by the so and so.(Link of the data that can be downloaded). From this data we have produced graphs for the number of cases and number of deaths that have caused due to all diseases in all the states of the US. From that we have extracted the data of the most dreadful disease that have caused maximum deaths and its activity during the years and from the plot it is easy to understand during which years the disease had its prime spread in the society. From that we can get which State has got higher damage due to the effect of this particular disease.




# 3.Methods

The R programming language gives us some quick and easy tools that let us convert our data into visually insightful elements like graphs. So that from the graphs the data becomes more interpretative and understandable.
The different types of graphs we have used in this work is listed below.


# 4.Data Analysis

The data analyses are done in different steps which are explained in detail as follows.

## 4.1 Loading the libraries

Here we can find all the libraries we have used in this work.


```{r}
library (ggplot2)
library (ggthemes)
library (readr)
library (tmap, mapview)
library (plotly)
library (viridis)
library (dplyr)
library (tidyverse)
library (ggpubr)
library (treemap)
library (usmap)
library (gapminder)
library (gganimate)
library (treemap)
library (ggridges)
library (png)
library (gifski)
options(scipen=999999) # It disables the scientific notion 
```
```{r}
options(scipen=9999)
```


```{r}
# setwd("K:/AI & DS/Data Visualisation/Tycho2") # Setting a directory for Project Work space. 

Tycho <- read.csv("K:/LSI/Life Science Info. SS 2022/Data Visulization/ProjectTycho_Level2_v1.1.0.csv", header = T, stringsAsFactors = T) # read.csv function loads a data in work space.

```

```{r}
dim(Tycho) # Dimension of the data in rows followed by columns
```

```{r}
Tycho <- Tycho %>% arrange(epi_week) #Arranging data by time sires from 1888 to 2014
head(Tycho)
```

```{r}
Tycho <- Tycho[ , -11] # Deletion of column num. 11 url
Tycho <- Tycho[ , -2] # Deletion of column num. 2 US
```

```{r}
Tycho <- Tycho %>% filter(number > 0 ) # removing 0 cases or deaths from the Tycho 36,59,360 - 26,78,605 = 9,80,755 rows has 0 cases data. 
```

```{r}
str(Tycho)# checking the data type of each column 
```

```{r}
Tycho$to_date <- as.Date(Tycho$to_date, format = "%Y-%m-%d") # change the format from factor to date
Tycho$from_date <- as.Date(Tycho$from_date, format = "%Y-%m-%d")
```


```{r}
levels(Tycho$disease) # List of all 50 disease
```

```{r}
unique(Tycho$state) # Here we have 57 observation in state but currently USA have 50 states only
```

```{r}
Disease <- Tycho %>% count(disease,event, wt = number) # sum of all disease according cases and deaths
```

Here is the another problem, for some disease data don't have records for cases, Only deaths are there for example "CHOLERA" and in reverse there is 0 death against huge cases of 4928414 for "CHLAMYDIA". Just for notice deaths of chlamydia in USA could be checked here, 
https://www.getargon.io/posts/health/conditions/us-std/death-rate-chlamydia-us/

```{r}
# Tree map to visualize disease ratio according total event per cases and deaths 
treemap(Disease, index = c("disease","event"),
        vSize = "n", vColor = "event", type = "index", bg.labels = 0,
        title = "  Treemap for all 50 Disease", border.col = c("black", "white"),
        border.lwds = c(0,0), palette = ("Set3"),
        align.labels = list(c("center","center"),c("right","bottom"))) 
```


Filtering cases and deaths on basis of States, Note some times in loc_type we have city name but its not in the most of cases so here consideration is all data as state data.


```{r}
Events <- Tycho %>% filter(number >= 1) # Separating values which shows at least one case or death.
Events$epi_week <- substr(Events$epi_week,1,4)
colnames(Events)[1] <- 'year' # Removing epi_week and making year column instead of year+ week.
Events$year <- as.numeric(Events$year)
```

Now let's understand the total cases for all periods in 50 states and with all diseases.

Below, the Geographic_map plot of US states shows the ratio of cases per state from 1888 to 2014. It also explains that some area or group of state presents darker colors and those states are primarily on the northwest side of the US, for instance, New York. In the south region, Texas and California in the west region presenting a dark color. In another case, states from central geo-location in the US present light colors which means less occurrences of cases during year 1888 to 2014. 

```{r}
Event_Case <- subset(Events, event == "CASES", select = c(year, state, disease, number))# Separating Cases from events column 
Event_Death <- subset(Events, event == "DEATHS", select = c(year, state, disease, number ))# # Separating Deaths from events column
```


```{r fig.align='center'}
# Preparing Data for Graph
year_state_Case <- Event_Case %>% select(year, state, number)
year_state_Case <- aggregate(year_state_Case, number ~ year + state, sum)

# Plotting a Graph according filtered Data
plot_geo(year_state_Case, locationmode = 'USA-states', frame = ~ year) %>% 
  add_trace(locations = ~ state, z = ~number, zmin = 0, zmax = max(year_state_Case$number), color = ~number, colorscale = 'PuBu') %>%
  layout(geo = list(scope = 'usa'), title = "State wise Cases event from all disease by each year ")

```

```{r fig.align='center'}
# Preparing Data for Graph
year_state_Death <- Event_Death %>% select(year, state, number)
year_state_Death <- aggregate(year_state_Death, number ~ year + state, sum)

# Plotting a Graph according filtered Data 
plot_geo(year_state_Death, locationmode = 'USA-states', frame = ~ year) %>% 
add_trace(locations = ~ state, z = ~number, zmin = 0, zmax = max(year_state_Death$number), color = ~number, colorscale = 'Electric') %>%
layout(geo = list(scope = 'usa'), title = "State wise Death event from all disease by each year ")
```


```{r}
# filtering disease which is grater than 50,000.
Disease1 <- Disease %>% pivot_wider(names_from = event, values_from = n, values_fill = 0) # creates Disease1 which has separate deaths and cases column 
Disease1 <- Disease1 %>% group_by(DEATHS,CASES) %>% filter(CASES  >= 55000 | DEATHS > 60000) # filtering cases and death
Disease1 <- Disease1 %>% pivot_longer( cols = DEATHS : CASES, names_to = "event", values_to = "number") # Changing back as before 

```

# Grouped Bar plot 1

The diseases with up to 50,000 documented cases are represented in this bar graph, along with the cases and deaths connected to each ailment. 25 diseases with up to 50,000 cases were found in the studies.Due to the fact that many diseases have fewer recorded cases than reported fatalities, the data is not entirely accurate.The shocking part is that there have been roughly 40,000 documented deaths from influenza and pneumonia combined when there were 0 reported cases.

```{r warning=FALSE, fig.align='center'} 
## Bar plot for comparing each disease with cases and deaths which are grater than 55,000
ggplot(Disease1, aes(x = reorder(disease, number) , y = reorder(number, disease),  fill= event))+
  geom_bar( position='dodge', stat='identity', ) + coord_flip()+ 
  scale_fill_manual( values = c("slategray3", "slategray4"),labels = c("CASES","DEATHS")) +  
  theme(text = element_text(size = 9),element_line(size =1),legend.position = c(-0.3, 0.0),axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Cases greater than 50,000") 
```

```{r}
# filtering disease which is lower than 50,000.
Disease2 <- Disease %>% pivot_wider(
  names_from = event, 
  values_from = n,
  values_fill = 0) # creates Disease1 which has separate deaths and cases column 
Disease2 <- Disease2 %>% group_by(DEATHS,CASES) %>% filter(CASES <= 50000 & disease != "PNEUMONIA AND INFLUENZA") # filtering cases and death and skip "PNEUMONIA AND INFLUENZA" from the selection because took already in Disease1.

Disease2 <- Disease2 %>% pivot_longer( cols = DEATHS : CASES, names_to = "event", values_to = "number")

```



```{r warning=FALSE, fig.align='center'} 
# Bar plot for comparing each disease with cases and deaths which are lower than 50,000
ggplot(Disease2, aes(x = reorder(disease, number) , y = reorder(number, disease), fill= event))+
  geom_bar( position='dodge', stat='identity') + coord_flip()+ 
  scale_fill_manual(values = c("lemonchiffon3", "lemonchiffon4"),labels = c("CASES","DEATHS")) +  
  theme(text = element_text(size = 10),element_line(size =1),legend.position = c(-0.4, 0.0),axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Cases lower than 50,000") 
```


Data shows only 15 disease from data which have data for deaths and cases, so if we look among those disease in box plot then it shows Tuberculosis, Pneumonia, pellagra and meningitis has high mortality as compare to other ten disease. 

```{r}
State_death_count <- filter (Tycho, event == "DEATHS") # filtering only deaths using state column (2)number from Tycho
State_death_count <- State_death_count %>%  group_by(state, disease) %>% summarise(number = sum(number))
```

```{r}
State_cases_count <- filter (Tycho, event == "CASES") # filtering only deaths using state column (2)number from Tycho
State_cases_count <- State_cases_count %>%  group_by(state, disease) %>% summarise(number = sum(number))

```

```{r}
State_death_count$groups <- cut(State_death_count$number, breaks = c(0,10,100,1000,10000,100000,500000)) # Creating Group of number of death for heat_map to create scale  
```


```{r fig.align='center'}
# Heat_map for deaths on the basis of ranges(0-10,10-100,100-1000...) which shows the concentration of the deaths per state.
ggplot(State_death_count, aes(state,disease)) +
  geom_tile(aes(fill= groups)) + # Border color
  scale_fill_manual(breaks = levels(State_death_count$groups), values = c("cornsilk","gold","orange","darkorange2","red2","darkred")) + # Box color low to high
  theme_tufte() + # plot theme
  theme(text = element_text(size = 9),axis.text.x = element_text(angle = 90, hjust = 1)) + # text size arrangement of scales
 theme(legend.key.width = unit(0.3, "cm"), legend.key.height = unit(0.2, "cm"),legend.direction = 'horizontal', legend.position = c(0.65, 1.05)) +
  ggtitle("Heatmap of Deaths in State" )
```
```{r}
State_cases_count$groups <- cut(State_cases_count$number, breaks = c(0,100,1000,10000,100000,1000000,5000000)) # Creating Group of number of cases for heat_map to create scale
```

```{r fig.align='center'}
# Heat_map for cases on the basis of ranges(0-100,100-1000,1000-10000...) which shows the concentration of the cases per state.
ggplot(State_cases_count, aes(state, disease )) +
  geom_tile(aes(fill= groups ) ) +
  scale_fill_manual(breaks = levels(State_cases_count$groups),  values = c("khaki","yellow3","olivedrab3","turquoise3","deepskyblue4","darkblue","mediumseagreen2")) +
  theme_tufte() + # plot theme
  theme(text = element_text(size = 7),axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(legend.key.width = unit(0.9, "cm"), legend.key.height = unit(0.1, "cm"),legend.direction = 'horizontal', legend.position = c(0.7, 1.05)) +
  ggtitle("Heatmap of Cases of Disease in State" ) # plot title
  
# White space shows there are no cases for particular disease for related states
```
```{r}
year_state_Case$Create <- cut(year_state_Case$number,  breaks = c(0,100,1000,10000,100000,1000000,5000000))
```

```{r fig.align='center'}
ggplot(year_state_Case, aes(x=year, y = state , fill= number))+ 
      geom_tile(aes(fill= Create)) +
      labs(title ='Heatmap of Cases on states per year', y = 'States') +
      scale_fill_manual(name = "Numbers", values = c("#FF9966","#FF3333","#CC0000","#990000","#660000", "#330000"), 
                        labels = c("^100", "^1000", "^10,000","^1,00,000","^10,00,000") ) +
      theme(text = element_text(size = 8))
```

```{r}
Both_Event <- Events %>% 
          select(year, state, disease, event, number)  %>% 
          pivot_wider(names_from = event, values_from = number , values_fn = {sum}) %>% 
          group_by(CASES, DEATHS) %>% 
          filter(CASES  > 0 & DEATHS > 0 ) %>%  
          pivot_longer( cols = DEATHS : CASES, names_to = "event", values_to = "number")
```



```{r fig.align='center'}
Box <- ggplot(Both_Event, aes(x = disease, y = number, fill = event)) +
  geom_boxplot( width = 0.6) + scale_y_log10(breaks = c(100,200,400,800,1600,3200)) +
  labs(x = "Disease", y = "Number") + scale_fill_manual(values= c("#666600", "#660000")) +
  ggtitle("Comparison of Deaths and Cases by Disease") +  
  theme(text = element_text(size = 8),axis.text.x = element_text(angle = 10, hjust = 1),legend.position = c(0.95, 0.9)) 
  
Box.animation = Box +
   transition_states(disease, wrap = FALSE) + shadow_mark(alpha = 0.5) +
   shadow_mark(alpha = 0.5) + enter_grow() + exit_fade() + ease_aes('back-out')
   
Box.animation
```


```{r}
# Here filtering the Measles according each decade,to combine it at the end.
Y1915 <- Tycho %>% filter(disease == "MEASLES") %>% group_by(state,number) %>% filter(from_date >= as.Date("1905-01-01")  & from_date < as.Date("1915-01-01"))  %>%  select(state, number) %>% ungroup(state,number) %>% group_by(state) %>% count(state, wt = number) %>% add_column(year = "1905") 

Y1925 <- Tycho %>% filter(disease == "MEASLES") %>% group_by(state,number) %>% filter(from_date >= as.Date("1915-01-01")  & from_date < as.Date("1925-01-01")) %>% select(state, number) %>% ungroup(state,number) %>% group_by(state) %>% count(state, wt = number) %>% add_column(year = "1915")

Y1935 <- Tycho %>% filter(disease == "MEASLES") %>% group_by(state,number) %>% filter(from_date >= as.Date("1925-01-01")  & from_date < as.Date("1935-01-01")) %>% select(state, number) %>% ungroup(state,number) %>% group_by(state) %>% count(state, wt = number) %>% add_column(year = "1925")

Y1945 <- Tycho %>% filter(disease == "MEASLES") %>% group_by(state,number) %>% filter(from_date >= as.Date("1935-01-01")  & from_date < as.Date("1945-01-01")) %>% select(state, number) %>% ungroup(state,number) %>% group_by(state) %>% count(state, wt = number) %>% add_column(year = "1935")

Y1955 <- Tycho %>% filter(disease == "MEASLES") %>% group_by(state,number) %>% filter(from_date >= as.Date("1945-01-01")  & from_date < as.Date("1955-01-01")) %>% select(state, number) %>% ungroup(state,number) %>% group_by(state) %>% count(state, wt = number) %>% add_column(year = "1945")

Y1965 <- Tycho %>% filter(disease == "MEASLES") %>% group_by(state,number) %>% filter(from_date >= as.Date("1955-01-01")  & from_date < as.Date("1965-01-01")) %>% select(state, number) %>% ungroup(state,number) %>% group_by(state) %>% count(state, wt = number) %>% add_column(year = "1955")

Y1975 <- Tycho %>% filter(disease == "MEASLES") %>% group_by(state,number) %>% filter(from_date >= as.Date("1965-01-01")  & from_date < as.Date("1975-01-01")) %>% select(state, number) %>% ungroup(state,number) %>% group_by(state) %>% count(state, wt = number) %>% add_column(year = "1965")

Y1985 <- Tycho %>% filter(disease == "MEASLES") %>% group_by(state,number) %>% filter(from_date >= as.Date("1975-01-01")  & from_date < as.Date("1985-01-01")) %>% select(state, number) %>% ungroup(state,number) %>% group_by(state) %>% count(state, wt = number) %>% add_column(year = "1975")

Y1995 <- Tycho %>% filter(disease == "MEASLES") %>% group_by(state,number) %>% filter(from_date >= as.Date("1985-01-01")  & from_date < as.Date("1995-01-01")) %>% select(state, number) %>% ungroup(state,number) %>% group_by(state) %>% count(state, wt = number) %>% add_column(year = "1985")

Y2005 <- Tycho %>% filter(disease == "MEASLES") %>% group_by(state,number) %>% filter(from_date >= as.Date("1995-01-01")  & from_date < as.Date("2005-01-01")) %>% select(state, number) %>% ungroup(state,number) %>% group_by(state) %>% count(state, wt = number) %>% add_column(year = "1995")
```

```{r}
# Combining the decade from 1915-2005 for measles disease
Measles_decade <- rbind(Y1915,Y1925,Y1935,Y1945,Y1955,Y1965,Y1975,Y1985,Y1995,Y2005) # Combining all decades to new data frame
Measles_decade$year <- as.integer(Measles_decade$year)
```


```{r fig.align='center'}
# Animation plot to catch the trend of disease per each state
ggplot(Measles_decade, aes( x = state, y = n, size = n, color = state)) +
  geom_point(alpha = 1.0, show.legend = FALSE  ) +
  scale_color_viridis_d() +
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(text = element_text(size = 9),axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(legend.position = "none") + scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}', x = 'State', y = 'Number of cases') + transition_time(year) 
```
# 5 Results
After this nominal analysis we found that there are several points to consider as a result. First fall data aren’t accurate or reliable for guarantied analysis because there are more than 50 states plus some disease have given only number of cases and some have have only deaths which is surly questionable that how figures of deaths noticed without figures of cases.

New York and Texas are highly affected states by cases as compare to other states. But in terms of mortality there are less impact on Texas compare to mortality rate in New York.

In terms of disease Measles have the highest cases approx 25309922, Also there isn’t any deaths recorded for Measles so we can’t create our hypothesis on the mortality impacts by Measles.

In addition there are several years haven’t recorded in Tycho project which is also a big factor to inhibits further analysis.

# 6 Discussion

The information provided by the US health department was practically defective and comprised state names, dates, and hours. In order to extract useful insights from the data, the numbers were further separated into events that discriminated between cases and fatalities as well as contained city and state names. The exact weekly time spans of the data allowed for analysis based on a series of occurrences.

The actual data shows no missing numbers, however a closer look reveals that the data lacks information on the death rates associated with a number of illnesses. just mention a few, there are chlamydia, gonorrhea, measles, mumps, and rubella. However, data on some illnesses only include fatalities. Consider the diseases varioloid, cholera, influenza and pneumonia.

When just cases were included, the research results showed that measles was the most common illness; however, if data on fatalities had been added, the technique may have yielded more accurate forecasts of mortality owing to measles.


# 7 Conclusion

This graphical study has shown that numerous illnesses, including meningitis, tuberculosis, pneumonia, and pellagra, have a high mortality rate in the United States, but the data shows that measles is the most prevalent disease among the 50.



